\begin{thebibliography}{}
\addcontentsline{toc}{chapter}{Bibliography}

\bibitem[Arbib and Manes 75]{arbib-manes-75} Arbib, M. A. and Manes, E.
G. (1975): {\it Arrows, Structures, and Functors --- The Categorical
Imperative ---}.  Academic Press.

\bibitem[Arbib and Manes 80]{arbib-manes-80} Arbib, M. A. and Manes, E.
G. (1980):  The Greatest Fixed Points Approach to Data Types.  In {\it
proceedings of Third Workshop Meeting on Categorical and Algebraic
Methods in Computer Science and System Theory}, Dortmund, West Germany.

\bibitem[Barr and Wells 85]{barr-wells-85} Barr, M. and Wells, C.
(1985): {\it A Series of Comprehensive Studies in Mathematics} Volume
278: {\it Toposes, Triples and Theories}.  Springer-Verlag.

\bibitem[Burstall and Goguen 77]{burstall-goguen-77} Burstall, R. M. and
Goguen, J. A. (1977): Putting Theories Together to Make Specification.
In {\it Proceedings of 5th International Joint Conference on Artificial
Intelligence}.  pp.\ 1045--1058.

\bibitem[Burstall and Goguen 80]{burstall-goguen-80} Burstall, R. M. and
Goguen, J. A. (1980): The Semantics of Clear: A Specification Language.
Internal Report CSR-65-80, Department of Computer Science, University of
Edinburgh.

\bibitem[Burstall and Goguen 81]{burstall-goguen-81} Burstall, R. M. and
Goguen, J.A. (1981): An Informal Introduction to Specifications using
Clear.  In {\it The Correctness Problem in Computer Sciences}, Academic
Press, pp.\ 185--213.

\bibitem[Burstall and Goguen 82]{burstall-goguen-82} Burstall, R. M. and
Goguen, J.A. (1982): {\it Algebras, Theories and Freeness: an
Introduction for Computer Scientists.}  Internal Report CSR-65-80,
Department of Computer Science, University of Edinburgh.

\bibitem[Burstall and Lampson 84]{burstall-lampson-84} Burstall, R. M.
and Lampson, B. (1984): A Kernel Language for Abstract Data Types and
Modules.  In {\it Lecture Notes in Computer Science}, Volume 173, pp.\
1--50.

\bibitem[Curien 86]{curien-86} Curien, P-L. (1986): {\it Categorical
Combinators, Sequential Algorithms and Functional Programming}.
Research Notes in Theoretical Computer Science, Pitman.

\bibitem[Dybjer 83]{dybjer-83} Dybjer, P. (1983): Category-Theoretic
logics and Algebras of Programs.  Ph.\ D. thesis, Department of Computer
Sciences, Chalmers University of Technology. University of Gothenburg.

\bibitem[Goguen and Burstall 83]{goguen-burstall-83} Goguen, J. A. and
Burstall, R. M. (1983): Introducing Institutions. In {\it proceedings of
Logic of Programming Workshop}.

\bibitem[Goguen and Tardo 79]{goguen-tardo-79} Goguen, J. A. and Tardo,
J. (1979): An Introduction to OBJ: A Language for Writing and Testing
Software Specifications.  In {\it Specification of Reliable Software},
IEEE, pp.\ 170--189.

\bibitem[Goguen, Thatcher and Wagner 78]{goguen-thatcher-wagner-78} Goguen,
J. A., Thatcher, J. W. and Wagner, E. G. (1978): An Initial Algebra
Approach to the Specification, Correctness and Implementation of
Abstract Data Types.  In {\it Current Trends in Programming
Methodology}, prentice-Hall, pp.\ 80--149.

\bibitem[Gordon, Milner and Wordsworth 79]{gordon-milner-wordsworth-79}
Gordon, M. J., Milner, A. J. and Wordsworth, C. P. (1979): {\it
Edinburgh LCF}.  {\it Lecture Notes in Computer Science}, Volume 78.

\bibitem[Goldblatt 79]{goldblatt-79} Goldblatt, R. (1979): {\it Topoi:
The Categorical Analysis of Logic}.  Studies in Logic and Foundation of
Mathematics, Volume 98, North-Holand.

\bibitem[Harper, MacQueen and Milner 86]{harper-macqueen-milner-86}
Harper, R., MacQueen, D. and Milner, R. (1986): {\it Standard ML}.  LFCS
Report Series, ECS-LFSC-86-2.  Department of Computer Science,
University of Edinburgh.

\bibitem[Kelly 72]{kelly-72} Kelly, G. M. (1972): Many-Variable
Functorial Calculus I.  In {\it Lecture Notes in Mathematics},
Volume 281, Springer-Verlag, pp.\ 66--105.

\bibitem[Lambek and Scott 86]{lambek-scott-86} Lambek, J. and Scott, P.
J. (1986): {\it Introduction to Higher-Order Categorical Logic}.  {\it
Cambridge Studies in Advanced Mathematics}, Volume 7.

\bibitem[Lawvere 63]{lawvere-63} Lawvere, F. W. (1963): Functorial
Semantics of Algebraic Theories.  In {\it Proceedings of the National
Academy of Science}, Volume 50, pp.\ 869--872.

\bibitem[Lehmann and Smyth 81]{lehmann-smyth-81} Lehmann, D. and Smyth,
M. (1981): Algebraic Specification of Data Types -- A Synthetic Approach
--.  {\it Mathematical System Theory}, Volume 14, pp.\ 97--139.

\bibitem[Mac Lane 71]{maclane-71} Mac Lane, S. (1971): {\it Categories
for the Working Mathematician}.  {\it Graduate Texts in Mathematics 5},
Springer-Verlag.

\bibitem[Manes 76]{manes-76} Manes, E. G. (1976): {\it Algebraic
Theories}.  {\it Graduate Texts in Mathematics 26}, Springer-Verlag.

\bibitem[Martin-L\"of 79]{martin-lof-79} Martin-L\"of, P. (1979):
Constructive Mathematics and Computer Programming.  Paper presented in
6th International Congress for Logic, Methodology and Philosophy of
Science.

\bibitem[Mendler 86]{mendler-86} Mendler N. P. (1986): {\it First- and
Second-Order Lambda Calculi with Recursive Types}.  Technical Report TR
86-764, Department of Computer Science, Cornell University.

\bibitem[Milner 84]{milner-84} Milner, R. (1984): {\it The Standard ML Core
Language}.  Internal Report CSR-168-84, Department of Computer Science,
University of Edinburgh.

\bibitem[Parasaya-Ghomi 82]{ghomi-82} Parasaya-Ghomi, K. (1982): Higher
Order Abstract Data Types.  Ph.\ D. Thesis, Department of Computer
Science, UCLA.

\bibitem[Plotkin 81]{plotkin-81} Plotkin, G. D. (1981): {\it A Structural
Approach to Operational Semantics}.  Technical Report DAIMI FN--19,
Computer Science Department, {\AA}rhus University.

\bibitem[Rydeheard 81]{rydeheard-81} Rydeheard, D. E. (1981):
Application of Category Theory to Programming and Program Specification.
Ph.\ D. thesis, University of Edinburgh.

\bibitem[Rydeheard and Burstall 86]{rydeheard-burstall-86} Rydeheard, D.
E. and Burstall, R. M. (1986): {\it Computational Category Theory}.

\bibitem[Scott 76]{scott-76} Scott, D. (1976): Data Types as Lattices.
{\it SIAM Journal of Computing}, Volume 5, pp.\ 552--587.

\bibitem[Smyth and Plotkin 82]{smyth-plotkin-82} Smyth, M. B. and
Plotkin, G. D. (1982): The Category-Theoretic Solution of Recursive
Domain Equations.  {\it SIAM Journal of Computing}, Volume 11.

\bibitem[Stenlund 72]{stenlund} Stenlund, S. (1972): {\it Combinators,
$\lambda$-Terms and Proof Theory}.  D. Reidel, Dordrecht.

\bibitem[Stoy 77]{stoy-77} Stoy, J. E. (1977): {\it The Scott-Strachey
Approach to Programming Language Theory}.  MIT Press.

\bibitem[Tait 67]{tait-67} Tait, W. (1967): Intentional Interpretation
of Functionals of Finite Type {I}.  {\it Journal of Symbolic Logic}, 32,
pp.\ 198--212.

\bibitem[Troelstra 73]{troelstra-73} Troelstra, A. S. (1973): {\it
Mathematical Investigation of Intuitionistic Arithmetic and Analysis}.
{\it Lecture Notes in Mathematics}, Volume 344, Springer-Verlag.
\end{thebibliography}
